Reconstructing the Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-11-14 doc/paper-lcr.tex
more draft editing

High precision cosmological distance measurements towards individual objects
such as time delay gravitational lenses or type Ia supernovae are affected by
weak lensing perturbations by galaxies and groups along the line of sight.
Such massive structures contribute an ``external convergence'' $\kappa$ that
in some cases dominates the uncertainty in the inferred distances.  We
characterise this uncertainty by marginalising over a probability density
function for $\kappa$: here, we investigate the use of a simple halo model to
estimate $\pr(\kappa|\data)$ given noisy estimates of the photometric redshift
and stellar mass of galaxies observed in optical imaging surveys of the field
in question. We use mock catalogs from the Millennium Simulation, and compare
our reconstructed $\pr(\kappa|\data)$ to ray-traced $\kappa$ truth values: 
taking into account realistic uncertainties on redshift and stellar masses, we
typically find that the reconstruction yields an improvement in precision of
40\% over the global $\pr(\kappa)$. For any given line of sight the
convergence estimate can be made before the investment of expensive follow-up
observations: we find that selecting systems by the precision with which
$\kappa$ can be estimated leads to a low convergence  sample that would give
time delay distance estimates a factor of 2 more precise than a random
selection,  with biases as low as 0.3\%. Preferentially selecting lines of
sight with high external shear (as might happen in an assembled sample of
quadruply-imaged time delay lenses) has the opposite effect, with distances
biased by around 1\%. We suggest areas for the improvement of this general
analysis framework that should allow yet more accurate cosmological inferences
to be made.

@@

\begin{keywords}
  gravitational lensing   --
  methods: statistical    --
  galaxies: halos         --
  gaaxies: mass function  --
  cosmology: observations
\end{keywords}

@@

In order to propogate the effect of line-of-sight perturbations into 
uncertainties on time-delay distances we require the probability density function (PDF) for the external convergence
given our available data $\data$, $\Pr(\xkappa|\data)$
% PHIL's REVIEWS OF THE LITRATURE
\comments{\citet{GunnarssonEtal2006} investigated a related PDF,  for the
magnification, using a galaxy halo model. Applying empirical galaxy
scaling relations both to simulate mock catalogs and then to reconstruct
the line of sight mass distribution. While they did not explore highly
overdense lines of sight, or the impact of groups and clusters, they
found that under their simplifying assumptions, the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gammax|\data)$ from their survey data, running similar Monte Carlo
reconstructions of the mass in nine lens fields based on
\citeauthor{WilliamsEtal2006} and \citeauthor{MomchevaEtal2006}'s
photometric and spectroscopic measurements of galaxies close to the line
of sight. They then  compared the resulting predicted shear
distributions with the external shear demanded by the strong lens model.
Their reconstructions showed most of the shear to have been generated by
bright galaxies within 2~arcminutes of the lens, and that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear. They found significant
discrepancies between the lens model shear and the Monte Carlo
predictions, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies.
%
\citet{KarpenkaEtal2012} used a halo model to predict deterministically 
the magnification due to
galaxies along the line of sight to type Ia supernovae, and inferred the
parameters of scaling relations between light and mass in these galaxies
from hundreds of supernova lightcone simultaneously. Following
\citet{JonssonEtal2010}, they subtract off the mean convergence to ensure that
the universal mass budget is balanced, and compare models where the halos are
all have either truncated singular isothermal profiles, or 
NFW profiles. Both groups were able to 
detect lensing effects in the supernova data this way.
}
% TOM's DISTILLING OF SAID REVIEWS.
Previous investigation of line-of-sight perturbations have focused on 
magnification and shear.\citet{GunnarssonEtal2006} used a galaxy halo model combined with simple scaling relations to reconstruct
the line of sight mass distribution and found that the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two. Conversely \citet{KarpenkaEtal2012} used the brightness dispersion in type Ia supernovae to infer the parameters of both their galaxy halo model and mass-to-light scaling relation. Both \citeauthor{KarpenkaEtal2012} and \citet{JonssonEtal2010} were able to 
detect lensing effects in the supernova data this way.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gammax|\data)$ from their survey data,
\citep{WilliamsEtal2006;MomchevaEtal2006} of galaxies close to the line
of sight. Comparing the predicted shear
distributions with the external shear demanded by the strong lens model 
they found significant
discrepancies, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies. \citeauthor{WongEtal2011} also found that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear.

Large numerical simulations have been used to
predict, from first principles, the distribution of likely external
shears and convergences along various lines of sight, including those
containing strong lens systems.  \citet{Holder+Schechter2003} and
\citet{DalalEtal2005} carried out ray-tracing calculations in N-body
simulations to estimate the distribution of external shear values. 
\citet{HilbertEtal2009} performed similar ray-tracing experiments
through the Millenium Simulation \citep{SpringelEtal2005}, generating a
predicted $\kappa$ at every position in a simulated sky.
\citet{HilbertEtal2009} found that \MS lines of sight with strong lenses
were not biased towards high $\kappa$, although selection functions for
discovering or using particular samples of strong lenses were not taken
into account.

The \MS results have been used to analyse real observations as well. 
\citet{SuyuEtal2010} selected \MS lines of sight by their apparent
galaxy overdensity, in a 45 arcsec radius aperture down to $i < 24.5$,
to match the observed overdensity towards the time delay lens
B1608$+$656 \citep{FassnachtEtal2011}.  The resulting distribution of
$\kappa$ values from the ray-tracing was taken to be an estimate of 
$\Pr(\kappa)$, which was then marginalised over when inferring the time
delay distance in this system. A similar proceedure was used in \citet{SuyuEtal2012b}.
% B1608$+$656 is indeed a high $\kappa$
% line of sight, but it does not reside in a massive cluster environment.
In a companion paper to this one, Greene et al. (submitted) investigate
improvements to this method by weighting the galaxy counts by galaxy
mass, and perpendicular distance to the line of sight, again using the
\MS mock catalogs and their associated $\kappa$ values to construct
$\Pr(\kappa|\data)$. For the most overdense lines of sight the
Green et al. results  show a $\sim$20\% improvement over
number counts alone, but little improvement for less dense lines.

In this paper, we combine the halo model reconstruction approach of
\citeauthor{GunnarssonEtal2006}, \citeauthor{JonssonEtal2010}, 
\citeauthor{WongEtal2011} and others, with the
idea of calibrating to simulations from \citeauthor{SuyuEtal2010}.
Taking the $\kappa$ values from
\citeauthor{HilbertEtal2009}'s ray-tracing calculations as the ``truth''
that we have to recover, we use the \MS mock galaxy catalogs first to
{\it calibrate} the reconstruction, and account for unseen mass and
voids, and then to {\it test} the accuracy of the line of sight mass
reconstruction under this assumption. Probabilistically assigning mass
and redshift to every observed galaxy in a given observed field, we
generate Monte Carlo sample line of sight mass distributions, and so
construct the PDF $\Pr(\kappa|\data)$ for that field. We then emulate
the combination of many such PDFs to quantify the residual biases that
would be translated to the global (including cosmological) parameters.
In doing so, we aim to answer the following questions:

@@

\item Faint galaxies, filaments and dark structures will not appear in
any photometric object catalog, but they will contribute convergence at
some level. How much of the total convergence comes from visible
galaxies, and how much effect do dark structures and voids have?

@@

\item We have some choices to make when planning follow-up observations
of interesting sightlines. How should we select objects to achieve the
highest accuracy convergence reconstruction? Can such a selection be
made robustly, using the reconstruction results?

@@

This paper is organized as follows. We review the relevant gravitational
lensing theory in~\Sref{sec:theory}, and the Millenium Simulation
ray-traced convergences and mock catalogs in \Sref{sec:MS}, before
introducing our simple reconstruction  model in \Sref{sec:model}. We
then test this model in two phases: first, in \Sref{sec:knownMh+z}, with
known redshift and halo mass for every galaxy in a lens field, in order
to quantify the irreducible uncertainty due to unseen mass, and second,
in \Sref{sec:obsMstar+z}, with realistic observational uncertainties on
the observed galaxies' stellar masses and redshifts. In
\Sref{sec:biases} we investigate the potential systematic error induced
by selecting a subset of sightlines. We discuss our results in
\Sref{sec:discuss} before concluding in \Sref{sec:conclude}.

@@

In this work, we have used 
the \MS as a calibration tool. If the real Universe is not
like the \MS, then this calibration will be incorrect. Repeating this 
study using
a different simulation to make the test catalogs would allow us to quantify
the sensitivity of our results to the simulation used. Similarly the use of a
different semi-analytic model to paint galaxies onto halos, and an alternative
stellar mass--halo mass relation would both enable similar exploration of
these systematic errors. It is possible that baryonic physics may alter
the radial profile of dark matter halos, even on large scales. \citet{Semboloni+2012}
found that baryonic physics had a 30-40\% impact on three-point shear statistics, it is
unclear how baryons affect one-point convergence statistics.

@@

\item  For the most overdense lines of sight, the reconstruction produces a
very broad PDF; conversely, we find that the lines of sight with the sharpest
inferred $\pr(\kappa|\data)$ are typically under-dense. Since the
reconstructions described here can be performed before follow-up time is
invested, it will be possible to  select targets by the precision of their
$\kappa$ estimates.

\item Photometric data is sufficient  to select the best lines of sight for
follow-up. Spectroscopic coverage helps to tighten  $\pr(\kappa|\data)$ for
high $\kappa$ lines of sight, but is of limited use for low $\kappa$ lines of
sight.

\item Selecting from the highest precision 33\% of the reconstructed lines of sight
results in a sample with low bias, 0.2\% in terms of the time delay distance,
and a statistical uncertainty due to line of sight mass structure that is half
that of the parent sample. Neglecting other sources of error, the residual systematic error would be equal to the
overall statistical uncertainty in a sample of around 30 lenses.

\item Conversely, selecting lines of sight with high shear ($|\gamma| >
0.03$ (as might be done inadvertently by focusing on four-image or wide
separation lenses) results in a sample with bias of 1\% in time delay
distance. Such a systematic error would become significant in a sample of $\sim$8 lenses.

>>>>

High precision cosmological distance measurements towards individual objects
such as time delay gravitational lenses or type Ia supernovae are affected by
weak lensing perturbations by galaxies and groups along the line of sight.
Such massive structures contribute an ``external convergence'' $\kappa$ that
in some cases dominates the uncertainty in the inferred distances.  We
characterise this uncertainty by marginalising over a probability density
function for $\kappa$: here, we investigate the use of a simple halo model to
estimate $\pr(\kappa|\data)$ given noisy estimates of the photometric redshift
and stellar mass of galaxies observed in optical imaging surveys of the field
in question. We use mock catalogues from the Millennium Simulation, and compare
our reconstructed $\pr(\kappa|\data)$ to ray-traced $\kappa$ truth values: 
taking into account realistic uncertainties on redshift and stellar masses, we
typically find that the reconstruction yields an improvement in precision of
40\% over the global $\pr(\kappa)$. For any given line of sight the
convergence estimate can be made before the investment of expensive follow-up
observations: we find that selecting systems by the precision with which
$\kappa$ can be estimated leads to a low convergence  sample that would give
time delay distance estimates a factor of 2 more precise than a random
selection,  with biases as low as 0.3\%. Preferentially selecting lines of
sight with high external shear (as might happen in an assembled sample of
quadruply-imaged time delay lenses) has the opposite effect, with distances
biased by around 1\%. We suggest areas for the improvement of this general
analysis framework that should allow yet more accurate cosmological inferences
to be made.

@@

\begin{keywords}
  gravitational lensing   --
  methods: statistical    --
  galaxies: halos         --
  galaxies: mass function  --
  cosmology: observations
\end{keywords}

@@

In order to propagate the effect of line-of-sight perturbations into 
uncertainties on time-delay distances we require the probability density function (PDF) for the external convergence
given our available data $\data$, $\Pr(\xkappa|\data)$
% PHIL's REVIEWS OF THE LITRATURE
\comments{\citet{GunnarssonEtal2006} investigated a related PDF,  for the
magnification, using a galaxy halo model. Applying empirical galaxy
scaling relations both to simulate mock catalogues and then to reconstruct
the line of sight mass distribution. While they did not explore highly
over-dense lines of sight, or the impact of groups and clusters, they
found that under their simplifying assumptions, the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gammax|\data)$ from their survey data, running similar Monte Carlo
reconstructions of the mass in nine lens fields based on
\citeauthor{WilliamsEtal2006} and \citeauthor{MomchevaEtal2006}'s
photometric and spectroscopic measurements of galaxies close to the line
of sight. They then  compared the resulting predicted shear
distributions with the external shear demanded by the strong lens model.
Their reconstructions showed most of the shear to have been generated by
bright galaxies within 2~arcminutes of the lens, and that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear. They found significant
discrepancies between the lens model shear and the Monte Carlo
predictions, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies.
%
\citet{KarpenkaEtal2012} used a halo model to predict deterministically 
the magnification due to
galaxies along the line of sight to type Ia supernovae, and inferred the
parameters of scaling relations between light and mass in these galaxies
from hundreds of supernova lightcone simultaneously. Following
\citet{JonssonEtal2010}, they subtract off the mean convergence to ensure that
the universal mass budget is balanced, and compare models where the halos are
all have either truncated singular isothermal profiles, or 
NFW profiles. Both groups were able to 
detect lensing effects in the supernova data this way.
}
% TOM's DISTILLING OF SAID REVIEWS.
Previous investigation of line-of-sight perturbations have focused on 
magnification and shear.\citet{GunnarssonEtal2006} used a galaxy halo model combined with simple scaling relations to reconstruct
the line of sight mass distribution and found that the dispersion in
apparent source brightness due to lensing magnification could be reduced
by a factor of two. Conversely \citet{KarpenkaEtal2012} used the brightness dispersion in type Ia supernovae to infer the parameters of both their galaxy halo model and mass-to-light scaling relation. Both \citeauthor{KarpenkaEtal2012} and \citet{JonssonEtal2010} were able to 
detect lensing effects in the supernova data this way.
%
\citet{WongEtal2011} estimated the PDF for the external shear,
$\Pr(\gammax|\data)$ from their survey data,
\citep{WilliamsEtal2006;MomchevaEtal2006} of galaxies close to the line
of sight. Comparing the predicted shear
distributions with the external shear demanded by the strong lens model 
they found significant
discrepancies, but were unable to distinguish between the environment
modelling and the strong lens modelling as the cause of these
discrepancies. \citeauthor{WongEtal2011} also found that both line of
sight structures and the group of galaxies in each lens plane contribute
significant proportions of the shear.

Large numerical simulations have been used to
predict, from first principles, the distribution of likely external
shears and convergences along various lines of sight, including those
containing strong lens systems.  \citet{Holder+Schechter2003} and
\citet{DalalEtal2005} carried out ray-tracing calculations in N-body
simulations to estimate the distribution of external shear values. 
\citet{HilbertEtal2009} performed similar ray-tracing experiments
through the Millennium Simulation \citep{SpringelEtal2005}, generating a
predicted $\kappa$ at every position in a simulated sky.
\citet{HilbertEtal2009} found that \MS lines of sight with strong lenses
were not biased towards high $\kappa$, although selection functions for
discovering or using particular samples of strong lenses were not taken
into account.

The \MS results have been used to analyse real observations as well. 
\citet{SuyuEtal2010} selected \MS lines of sight by their apparent
galaxy over-density, in a 45 arcsec radius aperture down to $i < 24.5$,
to match the observed over-density towards the time delay lens
B1608$+$656 \citep{FassnachtEtal2011}.  The resulting distribution of
$\kappa$ values from the ray-tracing was taken to be an estimate of 
$\Pr(\kappa)$, which was then marginalised over when inferring the time
delay distance in this system. A similar procedure was used in \citet{SuyuEtal2012b}.
% B1608$+$656 is indeed a high $\kappa$
% line of sight, but it does not reside in a massive cluster environment.
In a companion paper to this one, Greene et al. (submitted) investigate
improvements to this method by weighting the galaxy counts by galaxy
mass, and perpendicular distance to the line of sight, again using the
\MS mock catalogues and their associated $\kappa$ values to construct
$\Pr(\kappa|\data)$. For the most over-dense lines of sight the
Green et al. results  show a $\sim$20\% improvement over
number counts alone, but little improvement for less dense lines.

In this paper, we combine the halo model reconstruction approach of
\citeauthor{GunnarssonEtal2006}, \citeauthor{JonssonEtal2010}, 
\citeauthor{WongEtal2011} and others, with the
idea of calibrating to simulations from \citeauthor{SuyuEtal2010}.
Taking the $\kappa$ values from
\citeauthor{HilbertEtal2009}'s ray-tracing calculations as the ``truth''
that we have to recover, we use the \MS mock galaxy catalogues first to
{\it calibrate} the reconstruction, and account for unseen mass and
voids, and then to {\it test} the accuracy of the line of sight mass
reconstruction under this assumption. Probabilistically assigning mass
and redshift to every observed galaxy in a given observed field, we
generate Monte Carlo sample line of sight mass distributions, and so
construct the PDF $\Pr(\kappa|\data)$ for that field. We then emulate
the combination of many such PDFs to quantify the residual biases that
would be translated to the global (including cosmological) parameters.
In doing so, we aim to answer the following questions:

@@

\item Faint galaxies, filaments and dark structures will not appear in
any photometric object catalogue, but they will contribute convergence at
some level. How much of the total convergence comes from visible
galaxies, and how much effect do dark structures and voids have?

@@

\item We have some choices to make when planning follow-up observations
of interesting sight-lines. How should we select objects to achieve the
highest accuracy convergence reconstruction? Can such a selection be
made robustly, using the reconstruction results?

@@

This paper is organized as follows. We review the relevant gravitational
lensing theory in~\Sref{sec:theory}, and the Millennium Simulation
ray-traced convergences and mock catalogues in \Sref{sec:MS}, before
introducing our simple reconstruction  model in \Sref{sec:model}. We
then test this model in two phases: first, in \Sref{sec:knownMh+z}, with
known redshift and halo mass for every galaxy in a lens field, in order
to quantify the irreducible uncertainty due to unseen mass, and second,
in \Sref{sec:obsMstar+z}, with realistic observational uncertainties on
the observed galaxies' stellar masses and redshifts. In
\Sref{sec:biases} we investigate the potential systematic error induced
by selecting a subset of sightlines. We discuss our results in
\Sref{sec:discuss} before concluding in \Sref{sec:conclude}.

@@

In this work, we have used 
the \MS as a calibration tool. If the real Universe is not
like the \MS, then this calibration will be incorrect. Repeating this 
study using
a different simulation to make the test catalogs would allow us to quantify
the sensitivity of our results to the simulation used. Similarly the use of a
different semi-analytic model to paint galaxies onto halos, and an alternative
stellar mass--halo mass relation would both enable similar exploration of
these systematic errors. \citet{SuyuEtal2010} used the relative over-density to mitigate this
effect, but it does not totally remove the dependence on simulation.
It is also possible that baryonic physics may alter   
the radial profile of dark matter halos, even on large scales. \citet{Semboloni+2012}
found that baryonic physics had a 30-40\% impact on three-point shear statistics, it is
unclear how baryons affect one-point convergence statistics compared to the pure dark matter
simulations used in this work.

@@

\item  For the most over-dense lines of sight, the reconstruction produces a
broad PDF; conversely, we find that the lines of sight with the sharpest
inferred $\pr(\kappa|\data)$ are typically under-dense. Since the
reconstructions described here can be performed before follow-up time is
invested, it will be possible to select targets by the precision of their
$\kappa$ estimates. This will be useful
in an era when there are more known lenses than can be followed-up.

\item Photometric data is sufficient  to select the best lines of sight for
follow-up. Spectroscopic coverage helps to tighten  $\pr(\kappa|\data)$ for
high $\kappa$ lines of sight, but is of less use for low $\kappa$ lines of
sight.

\item Selecting from the highest precision 33\% of the reconstructed lines of sight
results in a sample with low bias, 0.1\% in terms of the time delay distance,
and a statistical uncertainty due to line of sight mass structure that is half
that of the parent sample. Neglecting other sources of error, the residual systematic error would be equal to the
overall statistical uncertainty in a sample of around 200 lenses.

\item Conversely, selecting lines of sight with high shear ($|\gamma| >
0.03$ (as might be done inadvertently by focusing on four-image or wide
separation lenses) results in a sample with bias of 0.3\% in time delay
distance. Such a systematic error would become significant in a sample of $\sim$30 lenses.
